Functional Pearl: Short and Mechanized Logical Relation for Dependent Type Theories

### 2024-12-16 paper/paper.tex
Edits of Sections 9.1, 9.2

\let\Bbbk\relax % amsmath tries to redefine this
\usepackage{amsthm,amssymb}
\let\draftnote\relax % this is already defined in JFP
\usepackage{draft}
\usepackage[outputdir=latex.out]{minted} % latexrun puts everything in latex.out
\usepackage{xspace,stmaryrd,mathtools,bbding}
\usepackage[para]{footmisc}
\usepackage{ottalt}
\inputott{rules}

@@

\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^{\Pi}$\xspace}
\newcommand{\CR}{\mathit{CR}}

@@

\begin{abstract}
Proof by logical relations is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relations in the
context of simply typed or polymorphic lambda calculi, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a countable universe
hierarchy, booleans, and an intensional identity type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

\section{Introduction}
This paper presents a \emph{short} and \emph{mechanized} proof of
logical consistency for \lang{}, a dependent type theory with a
predicative universe hierarchy, large elimination, an intensional identity
type, a boolean base type, and dependent elimination forms.

@@

The type $[[Set i]]$ represents a universe at level $i$,
a natural number. Abstractions $[[ \ x . a ]]$ and
dependent function types $[[Pi x : A . B]]$ are binding forms for the variable
$x$ in the body of the function and codomain of the function type.%
\footnote{In the exposition in this paper, binding forms are equal up to
  $\alpha$-conversion and we adopt the Barendregt Variable
  Convention~\cite{barendregt:lambda-calculus}, which lets us assume that
  bound variables are distinct.  In some places, we are informal about the
  treatment of variables and substitution; our mechanized proofs make these
  notions precise by using de Bruijn indices~\citep{debruijn1994automath}. }
We use the notation $[[A -> B]]$ when the output type $[[B]]$ is not dependent
on the input variable. Type annotations in abstraction forms are omitted, and
we discuss how the inclusion of type annotations can affect
our development in Section~\ref{sec:betaeta}, where we extend our
consistency result to the existence of $\beta\eta$-normal forms.
Intensional identity types $[[a ~ b : A]]$, or equality types, represent an
equality between terms $[[a]]$ and $[[b]]$ of the same type $[[A]]$, whose
canonical inhabitant is the reflexive proof $[[refl]]$. Equality proofs $[[p]]$
can be eliminated by the J eliminator $[[J c p]]$, which casts the term $[[c]]$
from one end of the equality to the other. Finally, we have booleans along with
the standard boolean values and conditional expressions.

@@

% \scw{This section pre-supposes that we want to define a logical relation,
% but doesn't precisely state what we want to use it to prove, and why a logical
% relation is suitable. (And why the property you want to prove is difficult to
% show!) Should add more motivation here.}
\scw{We need to explicitly point out that the key ideas of this paper are
  discussed, here, in this section.
  We need to explicitly remark on why logical relations are difficult to
  define for dependent type theory and explain why this setting is more
  difficult than with simple types (STLC) or with polymorphic types (System F).
  \begin{itemize}
  \item Large eliminations
  \item Definitional equality (not all types look like types)
  \end{itemize}
  Should we be more explicit in our comparison with Girard's trick for polymorphic type?
  There, the definition stays recursive because it doesn't substitute for the variables
  in the function types. But that approach is not available in this setting, because not
  all quantified things are types. And we might need that information to interpret, say,
  identity types in the right way.
}
\scw{ We also need to explicitly point out that our logical relation is untyped.
  This has two benefits: it allows semantic typing to be meaningful independent from
  syntactic typing (cite Derek, forward reference to next section) and it avoids
  significant bookkeeping, especially in the case of Kripke logical relations (we need to define
  what these are).
  Is there a cost to an untyped relation?
}

@@

\subsection{Logical relations for dependent types}
In the most general sense, a logical relation can be viewed as a
practical technique that uses a type-indexed relation to 
strengthen the induction hypothesis for the
property of interest. The original idea of this technique can be
traced back to
\citet{tait1967:reducibility}. This proof maps
types to sets of terms satisfying certain properties related to reduction.
The same idea is explained in \citet{girard1989proofs} and extended to
prove strong normalization of System F.
%
Tait's method has also been successfully applied to dependently typed
languages to prove strong normalization \cite{Martin-Lof-1973,luo1990extended,geuvers1994short, barendregt:lambda-calculi-with-types}.

However, the pen-and-paper representation of logical relations proofs
can be challenging to adapt to a theorem prover since many details
are hidden behind concise notations.
For example, \citet{geuvers1994short} presents the interpretation for types as
an inductively defined total function over the set of syntactically
well-formed types. \scw{The issue is that Geuver's metalogic is (untyped) set
theory, but Coq and Agda use type theory.}
In untyped set theory, it makes sense to define the logical relation as a
simply-typed function that takes a type and returns some set; however
in constructive type theory, the metalogic of Coq and Agda, the
interpretation function must be a dependently-typed function whose
return type depends on the derivation of the well-typedness of its
input. The well-typedness derivation and the proof of the
classification theorem\scw{what is the classification theorem}\yl{It
  says all well-formed terms A (either G |- A : .. or G |- .. : A)
then A is either an object, a type, or a kind. The classification is
used to guide erasure}
are examined in the body
of the interpretation function to decide whether an argument of an
application should be erased during interpretation. As a result, this
definition causes difficulties for modern proof assistants.
Due to the
impredicativity of the object language, \citet{geuvers1994short}'s
proof cannot be encoded in Agda, which has a predicative
metatheory. Due to the use of proof-relevant derivations, even in
Coq, a proof assistant that supports impredicativity, one would need
to constantly juggle between the impredicative but irrelevant sort
\texttt{Prop} sort and the predicative but relevant sort
\texttt{Type}.

More recent work such as \citet{Abel12}
and \citet{abel2008betaeta} make their definitions more explicit and
precise and thus more directly encodable in proof assistants.
% Our logical relation resembles their definition of a
% semantic universe hierarchy.
Unlike \lang, their systems support type-directed $\eta$ laws. These
type-directed equational rules complicate the proof significantly
since the confluence of full reduction is not available until the
fundamental theorem is proven. Therefore,
instead of a logical predicate, a relational model with
syntactic type annotations is needed to remove the early reliance on
confluence, despite the fact that the underlying type theory is
intensional.

\citet{coquand:canonicity,coquand:prop} use a \emph{proof-relevant} logical
relation to prove canonicity and normalization for intensional
dependent type theories. % Intuitively, proof-relevant logical relations
% can be viewed as the set theoretic augmented with information
% about syntactic terms so the fundamental theorem can give us more
Compared
to the proof-irrelevant logical predicate we present, Coquand's proof
has the benefit that it does not rely on the confluence of reduction,
and can be easily extended to more $\eta$ laws without relying on a
complex relational model. Instead of working with concrete syntax,
Coquand works directly with an algebraic representation of dependent
type theory called category with families (CwF)~\citep{hofmann:cwf},
though the idea of proof-relevant logical relations for dependent
types can be expressed independently of category
theory~\citep{barras2012semantical}.

More abstractly, the proof technique by Coquand can be viewed as an
instance of a categorical construction called Artin Gluing
~\citep{kaposi:gluing}. Different metatheoretic
properties, such as normalization, parametricity, and
$\Pi$-injectivity can be obtained by gluing the initial CwF with the
corresponding CwF model. \citet{sterling:tait} introduces the notion of
\emph{Synthetic Tait Computability}, which allows further
simplification of such proofs by working in the internal language
formed by gluing.
However, proof-relevant logical relations are more demanding on the
expressive power of the metatheory. For example, the mechanization of
NbE by \citet{altenkirch:normalisation} relies on Quotient Inductive
Inductive Types (QIIT)~\citep{kaposi:qiit}.

@@

%\newcommand\dheader[1]{\rotatebox{60}{{#1}}}
\newcommand\header[1]{\rotatebox{0}{{#1}}}

\begin{figure}[H]

% \begin{minipage}{0.8\textwidth}
  \begin{tabular}{| l |  c  | l | c | c | c | l | l | }
    \hline
%      & \header{Universes} & \header{Inductives} & \header{Conversion}
%      & \header{Large Elim} & Main results \\
      & \header{U} & \header{Ind} & \header{C} 
      & \header{L E} & \header{A} &  Main results \\

    \hline
    \lang{} (this work) & $\mathbb{N}$ & Id, Bool & U & $\checkmark$ & 1 
    & Consistency and normalization \\
    % \hline
    $\lambda^\theta$ & 0 & Id, Nat & U & $\XSolidBrush$ & 1 & Consistency \\
    Core Nuprl & $\mathbb{N}$ & W-Types & E & $\checkmark$ & 2 & Consistency \\
    NBE-in-Coq & 1 & Nat & T & $\XSolidBrush$ & 2 & Correctness of NBE \\
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidability of conversion \\
    MLTT-\`a-la-Coq & 1 & Id, Nat, $\Sigma$ & T & $\checkmark$ & 2 & Decidibility of type checking \\
    \hline
  \end{tabular}

  \begin{tabular}{ll}
    \\
  \underline{U}niverses: & Countable ($\mathbb{N}$), Zero (0), One (1) \\
  \underline{Ind}uctives: & Identity types (Id), Natural numbers (Nat), $\Sigma$-types ($\Sigma$), W-types \\
  \underline{C}onversion:& Untyped (U), Typed (T),  Extensional (E) \\
  \underline{L}arge \underline{E}liminations:& Included ($\checkmark$), not included ($\XSolidBrush$) \\
  \underline{A}rity of interpretation: & Sets of terms (1), Relations between terms (2) \\
  \\
  $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
  Core Nuprl &\citet{anand2014towards} \\
  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
  NBE-in-Coq& \citet{nbeincoq} \\
  MLTT-\'a-la-Coq &\citet{martin-lof-a-la-coq} \\
  \end{tabular}

@@

\scw{Should we add a column for whether their logical relation is
  unary or binary?}\yl{Technically a Kripke-style model has the
  context as an argument to the logrel, too. By binary, we are really
  talking about whether the interpreted set is binary or not. I think
  it is tricky to explain what we actually mean. }
\scw{I don't think it is that trick}
Figure~\ref{fig:featurematrix} presents several mechanized proofs
that feature logical-relations arguments for dependently-typed languages.
Each of these proofs is significantly larger than than our development; but they
also prove more results about different object languages.% \scw{Can we mention the
% line counts in the text, when they are known?}
The table provides a comparison between the various features of their
object languages, but is not exhaustive. For example,
\citet{casinghino:combining-proofs-programs} and
\citet{anand2014towards} both have support for partial
programs. However, we include features that we believe to be most
impactful to the definition of the logical relation.

\citet{casinghino:combining-proofs-programs} introduce $\lambda^\theta$, a dependently typed
programming language that uses modality to distinguish between logical
proofs and programs. % The programmatic fragment
% includes recursive data types and supports general recursion at the
% cost of introducing divergence.
% However, the $\lambda^\theta$
% language is limited in its expressiveness since it does not support
% type-level computation or polymorphism and therefore is not a fully dependently
% typed language according to \citet{abel2013normalization}.
The consistency proof of $\lambda^\theta$'s logical fragment has been
mechanized in Coq through a step-indexed logical relation;
step-indexing is required to model the programmatic fragment, which
interacts with the logical fragment.
The lack of polymorphism and type-level computation means their
logical relation can be defined recursively for well-formed types using
a size metric, which has been used in~\citet{liu2023dependently}.
Their development is around 8,000 lines of nonblank, noncomment code.

\citet{decagda} mechanize in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and a typed judgmental equality that includes the function
$\eta$ law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using the induction-recursion scheme,
which is available in Agda but not in Coq. Their development involves
around 10,000 lines of Agda code.
\citet{martin-lof-a-la-coq} transports the logical relation
from \citet{decagda} in the predicative fragment of Coq and further
extends the decidability of type conversion result
from~\citet{decagda} to the decidability type checking of a
bidirectional type system. Their development has around
30,000 lines of Coq code.

\citet{anand2014towards} mechanize the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. This metatheory is an
extensional type theory with features such as dependent functions,
inductive types, partial types, and a full universe hierarchy. They construct a PER
model in Coq to show the logical consistency of their language. Their
development has been further extended with features such as
intersection types, union types, and quotient types. The extensive
coverage of features results in a Coq development with around
330,000 lines of code.
\citet{nbeincoq}
mechanize the normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the
only base type from the object language, large elimination is not
supported despite the one predicative universe. Their development has
around 20,000 lines of Coq code.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage the
impredicative \texttt{Prop} sort of Coq to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further show it is possible
to encode a finite universe hierarchy without the use of
either impredicativity or induction-recursion. Their encoding of a countable
universe hierarchy relies on impredicativity, similar to our
development.

@@

\section{Conclusion}
\label{sec:conclusion}
In this work, we present a short and mechanized proof by logical relations for
a dependently typed language with a predicative universe hierarchy, large
elimination, an intensional identity type, and dependent eliminators.
From the fundamental theorem for the logical relation, we prove consistency
and canonicity, then show the extensibility of our approach by proving the
existence of $\beta$($\eta$)-normal forms and adding natural numbers and
cumulativity to the language. These extensions only require small and mechanical
changes to our proof development. Our Coq mechanization leverages existing Coq
libraries for reasoning about metatheory and for general purpose automation,
allowing us to significantly reduce the verbosity typically associated with
mechanized proofs. The result is a declarative proof style that rivals pen and paper.

>>>>

\let\Bbbk\relax % amsmath tries to redefine this
\usepackage{amsthm,amssymb}
\let\draftnote\relax % this is already defined in JFP
\usepackage{draft}
\usepackage[outputdir=latex.out]{minted} % latexrun puts everything in latex.out
\usepackage{xspace,stmaryrd,mathtools,pifont}
\usepackage[para]{footmisc}
\usepackage{ottalt}
\inputott{rules}

@@

\newcommand{\dotv}[2]{\href{#1}{\texttt{#1}}{\texttt{:#2}}}
\newcommand{\lang}{$\lambda^{\Pi}$\xspace}
\newcommand{\CR}{\mathit{CR}}
\newcommand{\cmark}{\ding{51}}
\newcommand{\xmark}{\ding{55}}

@@

\begin{abstract}
Proof by logical relations is a powerful technique that has been used
to derive metatheoretic properties of type systems, such as
consistency and parametricity. While there exists a
plethora of introductory materials about logical relations in the
context of simply typed or polymorphic lambda calculi, a streamlined
presentation of proof by logical relation for a dependently typed language
is lacking. In this paper, we present a short
consistency proof for a dependently typed language that contains a
rich set of features, including a countable universe
hierarchy, booleans, and a propositional equality type. We show that
the logical relation can be easily extended to prove the existence of
$\beta\eta$-normal forms.
We have
fully mechanized the consistency proof using the Coq proof assistant
in under 1000 lines of code, with 500 lines of additional code for the
$\beta\eta$-normal form extension.
\end{abstract}

@@

\section{Introduction}
This paper presents a \emph{short} and \emph{mechanized} proof of
logical consistency for \lang{}, a dependent type theory with a
predicative universe hierarchy, large elimination, a propositional equality
type, a boolean base type, and dependent elimination forms.

@@

The type $[[Set i]]$ represents a universe at level $i$,
a natural number. Abstractions $[[ \ x . a ]]$ and
dependent function types $[[Pi x : A . B]]$ are binding forms for the variable
$x$ in the body of the function and codomain of the function type.%
\footnote{In the exposition in this paper, binding forms are equal up to
  $\alpha$-conversion and we adopt the Barendregt Variable
  Convention~\cite{barendregt:lambda-calculus}, which lets us assume that
  bound variables are distinct.  In some places, we are informal about the
  treatment of variables and substitution; our mechanized proofs make these
  notions precise by using de Bruijn indices~\citep{debruijn1994automath}. }
We use the notation $[[A -> B]]$ when the output type $[[B]]$ is not dependent
on the input variable. Type annotations in abstraction forms are omitted, and
we discuss how the inclusion of type annotations can affect
our development in Section~\ref{sec:betaeta}, where we extend our
consistency result to the existence of $\beta\eta$-normal forms.
Propositional equality types $[[a ~ b : A]]$, represent an
equality between terms $[[a]]$ and $[[b]]$ of the same type $[[A]]$, whose
canonical inhabitant is the reflexive proof $[[refl]]$. Equality proofs $[[p]]$
can be eliminated by the J eliminator $[[J c p]]$, which casts the term $[[c]]$
from one end of the equality to the other. Finally, we have booleans along with
the standard boolean values and conditional expressions.

@@

% \scw{This section pre-supposes that we want to define a logical relation,
% but doesn't precisely state what we want to use it to prove, and why a logical
% relation is suitable. (And why the property you want to prove is difficult to
% show!) Should add more motivation here.}
\scw{We need to explicitly point out that the key ideas of this paper are
  discussed, here, in this section.
  We need to explicitly remark on why logical relations are difficult to
  define for dependent type theory and explain why this setting is more
  difficult than with simple types (STLC) or with polymorphic types (System F).
  \begin{itemize}
  \item Large eliminations
  \item Definitional equality (not all types look like types)
  \end{itemize}
  Should we be more explicit in our comparison with Girard's trick for polymorphic type?
  There, the definition stays recursive because it doesn't substitute for the variables
  in the function types. But that approach is not available in this setting, because not
  all quantified things are types. And we might need that information to interpret, say,
  equality types in the right way.
}
\scw{ We also need to explicitly point out that our logical relation is untyped.
  This has two benefits: it allows semantic typing to be meaningful independent from
  syntactic typing (cite Derek, forward reference to next section) and it avoids
  significant bookkeeping, especially in the case of Kripke logical relations (we need to define
  what these are).
  Is there a cost to an untyped relation?
}

@@

\subsection{Logical relations for dependent types}
In the most general sense, a logical relation is a
practical technique that uses a type-indexed relation to 
strengthen the induction hypothesis for the
property of interest. This technique originates from
\citet{tait1967:reducibility}, who maps
types to sets of terms satisfying certain properties related to reduction.
The same idea is explained by \citet{girard1989proofs} and extended to
prove strong normalization of System F.
Tait's method has also been successfully applied to dependently typed
languages to prove strong normalization \citep{Martin-Lof-1973,luo1990extended,geuvers1994short, barendregt:lambda-calculi-with-types}.

However, the pen-and-paper representation of logical relations proofs
can be challenging to adapt to a theorem prover since many details
are hidden behind concise notation.
For example, \citet{geuvers1994short} presents the interpretation for types as
an inductively defined total function over the set of syntactically
well-formed types.
In untyped set theory, it makes sense to define the logical relation as a
function that takes a type and returns a set; however,
in constructive type theory, the metalogic of Coq and Agda, the return type of
the interpretation function depends on the derivation of the well-typedness of
its input. The body of the interpretation function uses the well-typedness
derivation, along with a proof classifying whether the input is a kind, a type,
or a term, to determine whether the argument of an application should be erased
during interpretation.

As a result, Geuvers' interpretation causes difficulties for modern proof
assistants. Due to the impredicativity of the object language, the proof
cannot be encoded in Agda, whose metatheory is predicative. However, even
though Coq supports impredicativity, the proof relevance of the well-typedness
derivations would require juggling between the impredicative but irrelevant
\texttt{Prop} sort and the predicative but relevant \texttt{Type} sort.

More recent work such by \citet{Abel12} and \citet{abel2008betaeta} makes the
definitions more explicit and precise and thus more directly encodable in proof
assistants. Unlike \lang, their systems support type-directed $\eta$ laws.
These type-directed equational rules complicate the proof significantly
since the confluence of full reduction is not available until after the
fundamental theorem is proven. Therefore, instead of a logical predicate,
they need a relational model with syntactic type annotations to remove the early
reliance on confluence, in spite of the object type theory being intensional.

\citet{coquand:canonicity,coquand:prop} uses a \emph{proof-relevant} logical
relation to prove canonicity and normalization for intensional
dependent type theories. In contrast to the proof-irrelevant logical predicate
we present, Coquand's proof does not rely on the confluence of reduction,
and can be easily extended to more $\eta$ laws without relying on a
complex relational model. Instead of working with concrete syntax,
Coquand works directly with an algebraic representation of dependent
type theory using categories with families (CwF)~\citep{hofmann:cwf},
though the idea of proof-relevant logical relations for dependent
types can be expressed independently of category
theory~\citep{barras2012semantical}.

More abstractly, Coquand's proof technique is an instance of a categorical
construction called Artin gluing~\citep{kaposi:gluing}. Different
metatheoretic properties, such as normalization, parametricity, and
$\Pi$-injectivity, can be obtained by gluing the initial CwF with the
corresponding CwF model. \citet{sterling:tait} introduces the notion of
synthetic Tait computability, which allows further
simplification of such proofs by working in the internal language
formed by gluing.
However, proof-relevant logical relations are more demanding on the
expressive power of the metatheory. For example, the mechanization of
NbE by \citet{altenkirch:normalisation} relies on quotient
inductive--inductive types (QIIT)~\citep{kaposi:qiit}.

@@

\begin{figure}
  \begin{tabular}{| l |  c  | l | c | c | c | l | l | r | }
    \hline
%     \newcommand\header[1]{#1}
%     & \header{Universes} & \header{Inductives} & \header{Conversion}
%     & \header{Large Elim} & Main results \\
      & U & Ind & C & LE & A & Main results & LOC \\
    \hline
    \lang{} (this work) & $\mathbb{N}$ & Eq, Bool & U & \cmark & 1 
    & Consistency, normalization & 1331 \\
    $\lambda^\theta$ & 0 & Eq, Nat & U & \xmark & 1 & Consistency & \textasciitilde{}8k\\
    Core Nuprl & $\mathbb{N}$ & W types & E & \cmark & 2 & Consistency & \textasciitilde{}330k \\
    NBE-in-Coq & 1 & Nat & T & \xmark & 2 & Correctness of NbE & \textasciitilde{}20k \\
    $\lambda^{\Pi U\mathbb{N}}$ & 1 & Nat, $\Sigma$ & T & \cmark & 2 & Decidability: conversion & \textasciitilde{}10k \\
    MLTT-\`a-la-Coq & 1 & Eq, Nat, $\Sigma$ & T & \cmark & 2 & Decidibility: type checking & \textasciitilde{}30k \\
    \hline
  \end{tabular}

  \begin{tabular}{ll} \\
  \underline{U}niverses: & Countable ($\mathbb{N}$), zero (0), one (1) \\
  \underline{Ind}uctives: & Equality types (Eq), naturals (Nat), dependent pairs ($\Sigma$), W types \\
  \underline{C}onversion: & Untyped (U), typed (T),  extensional (E) \\
  \underline{L}arge \underline{E}limination: & Included (\cmark), not included (\xmark) \\
  \underline{A}rity of interpretation: & Sets of terms (1), relations between terms (2) \\
  \\
  $\lambda^\theta$  & \citet{casinghino:combining-proofs-programs} (logical fragment only) \\
  Core Nuprl &\citet{anand2014towards} \\
  $\lambda^{\Pi U\mathbb{N}}$ &\citet{decagda} \\
  NBE-in-Coq& \citet{nbeincoq} \\
  MLTT-\`a-la-Coq &\citet{martin-lof-a-la-coq} \\
  \end{tabular}

@@

Figure~\ref{fig:featurematrix} presents several mechanized proofs
that feature logical relations for dependently-typed languages.
Each of these proofs is significantly larger than our development, but they
also prove more results about different object languages.
The table provides a comparison between the various features of their
object languages, but is not exhaustive. For example,
\citet{casinghino:combining-proofs-programs} and
\citet{anand2014towards} both have support for partial
programs. However, we include features that we believe to be most
impactful to the definition of the logical relation.
For a basic comparison of development size, we include the approximate
nonblank, noncomment lines of proof code for each project.

\citet{casinghino:combining-proofs-programs} introduce $\lambda^\theta$,
a dependently typed programming language that uses modalities to distinguish
logical proofs from programs.
% The programmatic fragment
% includes recursive data types and supports general recursion at the
% cost of introducing divergence.
% However, the $\lambda^\theta$
% language is limited in its expressiveness since it does not support
% type-level computation or polymorphism and therefore is not a fully dependently
% typed language according to \citet{abel2013normalization}.
The consistency proof of $\lambda^\theta$'s logical fragment has been
mechanized in Coq using a step-indexed logical relation;
step-indexing is required to model the programmatic fragment, which
interacts with the logical fragment.
The lack of polymorphism and type-level computation means their
logical relation can be defined recursively for well-formed types using
a size metric, which has been used in~\citet{liu2023dependently}.

\citet{decagda} mechanize in Agda the decidability of type
conversion rule for a dependently typed language with one predicative
universe level and a typed judgmental equality that includes the function
$\eta$ law. They
use a Kripke-style logical relation parameterized over a
type-directed equivalence relation satisfying certain
properties to facilitate the reuse of their definition. The
logical relation is defined using induction--recursion,
which is available in Agda but not in Coq. Their development involves
around 10,000 lines of Agda code.
\citet{martin-lof-a-la-coq} adapts the logical relation
by \citet{decagda} to the predicative fragment of Coq and further
extends the decidability of type conversion result to the decidability of
type checking a bidirectional type system.

\citet{anand2014towards} mechanize the metatheory of
Nuprl~\citep{constable1986implementing} in Coq. The object theory is an
extensional type theory with dependent functions, inductive types,
partial types, and a full universe hierarchy. They construct a PER
model in Coq to show the logical consistency of the language. Their
development has been further extended with intersection types,
union types, and quotient types, resulting in a Coq development of around
330,000 lines of code.
\citet{nbeincoq} mechanize a normalization-by-evaluation algorithm in Coq for a
dependently typed language with one predicative universe, similar to
\citet{decagda} and \citet{martin-lof-a-la-coq}. However, since their
type system has no elimination form for natural numbers, the only base type,
large elimination is not supported.
Both \citet{anand2014towards} and \citet{nbeincoq} leverage
Coq's impredicative \texttt{Prop} sort to define the interpretation
of dependent function types and thus are closely related to our
mechanization. \citet{anand2014towards} further show it possible
to encode a finite universe hierarchy using neither impredicativity
nor induction--recursion. Their encoding of a countable
universe hierarchy relies on impredicativity, similar to our
development.
\jc{Wait I'm confused, do they use impredicativity or not?}

@@

\section{Conclusion}
\label{sec:conclusion}
In this work, we present a short and mechanized proof by logical relations for
a dependently typed language with a predicative universe hierarchy, large
elimination, a propositional equality type, and dependent eliminators.
From the fundamental theorem for the logical relation, we prove consistency
and canonicity, then show the extensibility of our approach by proving the
existence of $\beta$($\eta$)-normal forms and adding natural numbers and
cumulativity to the language. These extensions only require small and mechanical
changes to our proof development. Our Coq mechanization leverages existing Coq
libraries for reasoning about metatheory and for general purpose automation,
allowing us to significantly reduce the verbosity typically associated with
mechanized proofs. The result is a declarative proof style that rivals pen and paper.
